Stack Safety as a Security Property

### 2021-03-22 CheriApproaches.tex
Finished up some of the Cheri work I was doing

There are three special capability registers. The {\em default data capability} ({\tt \$ddc}),
is applied to reads and writes through non-capability pointers. The
{\em program counter capability} ({\tt \$pcc}) is applied to instruction fetches.
The {\em stack capability} ({\tt spc}) takes the place of the stack pointer and is used by
compiler generated code. Basic capabilities are supported by the following instructions:

\begin{tabular}{l l l}
  {\tt st}      & Data Store       & Stores data, clearing any capability bits \\
  {\tt ld}      & Data Load        & Loads data, clearing any capability bits \\
  {\tt cst}     & Capability Store & Stores a capability, but failstops rather than store
                                     a local one outside the stack \\
  {\tt cld}     & Capability Load  & Loads a capability from memory \\
  {\tt incbase} & Increment Base   & Increases the base of a capability by one and
                                     decrements the offset \\
  {\tt decbnd}  & Decrement Bound  & Decreases the bound of a capability by one \\
\end{tabular}

@@

\begin{tabular}{l l l}
  {\tt so} & Seal Object   & Takes a pair of capabilities and
                                  seals them with matching identifiers \\
  {\tt io} & Invoke Object & Takes matching sealed capabilities \(c\) and
                                  \(p\), jumps to \(p\), and unseals \(c\) \\
\end{tabular}

Finally, we will use monotone capabilities.

\subsection{The calling convention}

From here on out, assume all capabilities are local. They can live in registers and stack
memory only. In the interest of corresponding to out model, the {\tt \$ddc} is permissive
and allows access to everything that isn't stack or code, and the {\tt \$pcc} allows access
only to our code.

{\bf Initially:} The {\tt \$stc} encompasses the full stack in its bounds. There is a
special bit of code that has no function identifier that serves as a landing pad for
functions, that takes a capability on a two-word block, loads the first word to {\tt \$stc}
and jumps to the second. The jump is annotated as a return by the compiler.

{\bf Caller:} To make a call, the caller allocates two extra words on the stack. It copies
the current {\tt \$stc} and the address of the instruction after the call into these. Then
it creates a capability restricted to this object and links it as an object capability with
the address of the landing pad. The object capability is stored in a register. Then the
caller increments the base of the {\tt \$stc} until its offset is zero, which will prevent
access to its frame. Finally, the caller jumps to the call target -- this instruction is
annotated as a call by the compiler.

After a return, the caller immediately clears all non-return-value registers and the stack
above the {\tt \$stc}'s offset. This prevents a copy of the object capability from being
stashed somewhere for a multiple return attack.

{\bf Callee:} On entry, the callee allocates memory by increasing the {\tt \$stc} and stores
the object capability it was passed. To return, it invokes that capability, executing the
landing pad and unsealing the data portion of the object capability. At the end of the
landing pad sequence, there is a jump instruction, which is annotated as a return. As
discussed above, to dynamically determine the target of the return, we must look at the
machine state and contet: the data portion of the return capability points to the activation
record of the caller, which is labeled {\it Passed(d)} for some depth \(d\) in the context.
(That is, it lies with the frame of the activation at depth \(d\).) The return target is
therefore \(d\). An intuitive way of describing this is that on return, we unseal layers of
calls until the activation record we're using to return has been unsealed.

\subsection{Analysis}

This calling convention prevents the callee from accessing the caller's state. The caller
guarantees that there is no additional object capability floating around somewhere that
the callee could access, and the {\tt \$stc} has its bounds shrunk to exclude the caller's
frame. It can only be restored with the object invocation.

What it doesn't necessarily do is prevent the callee from reading its frame. An adversarial
caller could write above its stack pointer before the call and there would be no protection.
So we have ``sealed confidentiality'' instead of normal confidentiality that includes all
uninitialized memory.

Luckily, if we extend our model with the ``uninitialized capabilities'' of Georges et. al.
we can once again enforce the confidentiality of uninitializes memory, as well. Then
we have full local state encapsulation.

The generic return code being shared between all functions is a little unfortunate, as it
puts a hole in control separation. We'll need to describe the correct behavior carefully
and adjust the property.

We may have a problem with returns, in that a callee could potentially pass along its
return object to a further callee, which could then bypass the first and return straight
to the original caller. This appears to be a hazard of our attacker model.

The caller would be happy enough, as its state would have been protected,
but it messes up our whole model of depths and counting
of returns. And the callee might rely on getting returned to.

\section{Well-formedness, Reasonableness, and PIPE}

As discussed above, Skortengaard et. al. impose static well-formedness conditions on
all of their code, as well as certain

>>>>

There are three special capability registers. The {\em default data capability} ({\tt \$ddc}),
is applied to reads and writes through non-capability pointers. The
{\em program counter capability} ({\tt \$pcc}) is applied to instruction fetches.
The {\em stack capability} ({\tt \$stc}) takes the place of the stack pointer and is used by
compiler generated code. Basic capabilities are supported by the following instructions:

\begin{tabular}{l l l}
  {\tt st}      & Data Store       & Stores data, clearing any capability bits \\
  {\tt ld}      & Data Load        & Loads data, clearing any capability bits \\
  {\tt stc}     & Capability Store & Stores a capability, but failstops rather than store
                                     a local one outside the stack \\
  {\tt ldc}     & Capability Load  & Loads a capability from memory \\
%  {\tt incbase} & Increment Base   & Increases the base of a capability by one and
%                                     decrements the offset \\
%  {\tt decbnd}  & Decrement Bound  & Decreases the bound of a capability by one \\
\end{tabular}

@@

\begin{tabular}{l l l}
  {\tt seal}   & Seal Object    & Takes a pair of capabilities and a seal identifier and
                                  seals them with the identifier \\
  {\tt invoke} & Invoke Object  & Takes matching sealed capabilities \(c\) and
                                  \(p\), jumps to \(p\), and unseals \(c\) \\
\end{tabular}

Finally, we will use {\em uninitialized capabilities}. These have a special permission structure
where their offset only moves when written to. The portion of the capability below the offset,
which is initialized, can be promoted into a normal capability. They support the following
instructions:

\begin{tabular}{l l l}
  {\tt stu}  & Data Store (Uninitialized) & Stores data and increments offset of Ucap \\
  {\tt stuc} & Capability Store           & As {\tt stU} but stores a capability \\
  {\tt prom} & Promote                    & Promote initialized part of an uninitialized capability
                                            into a normal capability \\
  {\tt morp} & Co-promote                 & Remove the initialized part of an uninitialized capability \\
\end{tabular}

\section{A Capability Machine}

Here I'll describe a simple machine implementing the operations above. Eventually. For now
I mocked it up in CheriModel.v. It's pretty close to what's described above. The most interesting
thing is the semantics of invoking a sealed object. A sealed object is a pair of a code capability
and a data capability (though these are distinguished only by convention.) There is a special register,
{\tt \$invdst}, in which the unsealed data capability is stored (and of course, the unsealed code capability
is copied to the {\tt \$pcc}. The idea is that on entry to an invoked capability, we always need to know
which register has the data capability, so it can't be dynamic.

\subsection{A call}

Here we will give an example of a call to a function at address {\tt 20} that takes an
argument in register {\tt ra}. Let our call start at address {\tt 0}. Suppose that we have a piece
of code at {\tt 50} that executes a ``landing pad'' for a return. Assume that all capabilities referenced
here are local, and the stack capability is an uninitialized capability that can write locally. There
could be nonlocal capabilities as well, for use in the heap, but to match our model we can also just assume
that such capabilities are derived from a permissive {\tt $ddc}.

\begin{verbatim}
<0: [clear all registers]
0: mov 42 ra
1: mov $stc r1
2: morp $stc
3: stuc r1 $stc               [PASS $stc]
4: stuc [$pcc + 7] $stc       [PASS $stc]
5: mov $stc r1
6: prom r1
7: morp $stc
8: mov [$pcc + 42] r2
9: seal s r1 r2
10: jmp +10                   [CALL]
11: ...
\end{verbatim}

We can walk through what this does step by step. First, to make sure that we don't have any extra
capabilities laying around, we clear all of our registers. Then we set the argument in {\tt ra}.
Now things get interesting: instructions 1-2 save the current stack capability and cut off
the readable portion, leaving only the uninitialized portion. Instructions 3-4 construct an activation
record containing the previous stack capability and the appropriate return address. Note that each {\tt stuc}
instruction increments the stack pointer, so we are implicitly allocating these two variables.
They will need to be accessible to the callee, so we annotate them as passed. Next, we create a capability
that can only access them, in instructions 5-6: the promoted stack capability is a capability on
only the written portion. Finally instruction 7 ensures that the stack capability cannot access
them. Instructions 8 and 9 now seal together the data capability and a code capability on address
50, where our ``landing pad'' code lives, and address 10 jumps us ahead to our function entry.

Now, what happens when we return? Let's look at the landing pad. Recall that we have a special
register {\tt \$invdst} that is where the unsealed data capability lives.

\begin{verbatim}
50: ldc $invdst $stk
51: jalr [$invdst+1]          [RETURN]
\end{verbatim}

Pretty simple! Notice something interesting: if we invoke a (properly constructed) return capability,
we {\it always} hit a {\tt jalr} labeled as a return two steps later. In other words, an adversary can't do
something that looks like a return without it being labeled a return.

Now it's pretty easy here to argue some of our properties. We should have that no unsealed capabilities are
present unless they were passed as arguments, except the stack capability, which is completely uninitialized.
So we should not be able to read anything without writing it or access a caller's frame at all.

When we've returned, we've restored the old stack capability, so the callee's frame is again uninitialized.
But! We may have a capability on it that the callee returned. That's a potential problem, because the capability
could then be passed to a future call in a way that allows it to persist. There's a solution to that, unpleasant
though it is: clear every register on return except for the value being returned, and manually check that it
isn't a capability that overlaps the uninitialize portion of the stack. Or indeed, any stack capability at all.

\subsection{Friction points}

In this model we don't really have control separation. And if we did, the landing pad wouldn't
actually belong to any one function, so it'd be a bit odd. We could get separation back by making
the call an invocation of an object, with its own launchpad that handles all of the rest of the
construction.

We still have the problem with returns, in that a callee could potentially pass along its
return object as an argument to a further callee, which could then bypass the first and return
straight to the original caller. We can model what happens in this case by checking, at the return point,
where we are in the stack - that is the depth we are about to return to, and it may be more than one depth
down (or zero, as the callee could always manufacture its own return object to ``return'' to itself).
